---
title: "RA Synovial/Blood continuous latent factors"
author: "Andrew McDavid"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
  slidy_presentation:
    font_adjustment: +1
extensions: inline_footnotes
---

# Libraries

```{r library, message = FALSE, warning = FALSE, results = 'hide'}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, autodep=TRUE, message=FALSE, warning=FALSE)
knitr::opts_chunk$set(dev = c('png', 'pdf'))
library(broom)
library(ggbeeswarm)
library(ggrepel)
library(SingleCellExperiment)
library(scater)
library(slingshot)
library(tidyverse)
library(dplyr)

## Other packages from bioconductor:
# DropletUtils, scran.  Install with BiocInstaller::biocLite(c('DropletUtils', 'scran'))
## Other packages from cran:
# install.packages(c('Seurat', 'readxl'))

```

# Load data

```{r}
cdata = read_csv('refined/filtered_clustered_cdata.csv', guess_max = 1e4) %>% mutate(res.0.5 = factor(res.0.5))
fdata = read_csv('refined/filtered_clustered_fdata.csv', guess_max = 5e5) %>% select(-matches('rank|total|mean|bio|tech|p.value|FDR|hvg'))
counts = readRDS('refined/filtered_clustered_exprs.rds')

gene_gradlist = readxl::read_xlsx('refined/curated gene list for gradient analysis 052920.xlsx', col_names = c('symbol', 'X__2', 'comment')) %>% mutate(curated = TRUE) %>% select(-comment, -X__2)
gene_gradlist = gene_gradlist[!duplicated(gene_gradlist$symbol),]
fdata = left_join(fdata, gene_gradlist, by = 'symbol')

sce = SingleCellExperiment(assays = list(counts = counts), rowData = fdata, colData = cdata)

```
# Normalize

```{r sizeFactors}
qc = scran::quickCluster(sce, method = 'igraph')
sce = scran::computeSumFactors(sce, cluster = qc)
summary(sizeFactors(sce))
sum(sizeFactors(sce)<0)
```

# Tbet groups (old analysis)
```{r setupTests}

colData(sce) = cbind(colData(sce), assay(sce)[!is.na(rowData(sce)$group),] %>% as.matrix %>% t)
tbx = factor(colData(sce)$TBX21>0, labels = c('TBX21-', 'TBX21+'))
itgax = factor(colData(sce)$ITGAX>0, labels = c('ITGAX-', 'ITGAX+'))
colData(sce)$grp = interaction(tbx, itgax)

props = colData(sce) %>% as.data.frame() %>% group_by(dataset, res.0.5, grp) %>% summarize(count = n())

props = props %>% summarize(total = sum(count)) %>% left_join(props) %>% mutate(prop = count/total)


plt = ggplot(props, aes(x = dataset, fill = grp))+geom_col(aes(y = prop)) + theme_minimal()+  scale_fill_brewer(type = 'qual', palette = 'Set1') + coord_flip()+ facet_wrap(~res.0.5)
plt

```


```{r normalize_sf}
qc = qc[sizeFactors(sce)>0]
sce = normalise(sce[,sizeFactors(sce)>0])
sce$IGHDhi = factor(kmeans(logcounts(sce)['IGHD',], centers = c(0, 3))$cluster)
plt = qplot(x = sizeFactors(sce), y = Matrix::colSums(assay(sce, 'counts')), color = factor(qc)) + geom_smooth() 
plt + ylab('Library size (sum)')
plt + aes(y = Matrix::colSums(assay(sce, 'counts') > 0)) + ylab('CDR')

```

# Cloupe

```{r}
cloupein = read_csv('refined/Cloupe-barcodes.csv')
aggr_in = read_csv('refined/aggregation.csv') %>% mutate(suffix = seq_len(nrow(.)))
sce_remap = colData(sce) %>% as.data.frame %>% mutate(library_id = dataset)
source('Interface-Loupe.R')
sce_cloupe = remap_cloupe(cloupin, aggr_in, sce_remap)

```

```{r eval = TRUE}
cloupeout1 = left_join(cloupein, sce_cloupe) %>% select(Barcode, res.0.5)
write_csv(cloupeout1, 'refined/filtered_clustered_only_clusterid.csv')

cloupeout2 = right_join(cloupein, sce_cloupe) %>% select(Barcode, tSNE_1, tSNE_2)
write_csv(cloupeout2, 'refined/filtered_clustered_only_tsne.csv')
```

### Write BCR info

```{r, eval = TRUE}
get_barcode = . %>% mutate(Barcode_prefix = str_replace(barcode, '-1', '')) %>%
    right_join(sce_cloupe %>% select(Barcode, dataset, Barcode_prefix), by = c('Barcode_prefix', 'dataset'))

cloupeout3 = left_join(cloupein, sce_cloupe) %>% select(Barcode, chain_type)
write_csv(cloupeout3, 'refined/filtered_clustered_bcr_chaintype.csv')

cloupeout4 = sce_cloupe %>% mutate_at(vars(class_heavy, class_light), fct_explicit_na) %>% mutate(paired_class = str_c(class_heavy, "-", class_light)) %>%
  right_join(cloupein) %>% select(Barcode, paired_class)
write_csv(cloupeout4, 'refined/filtered_cluster_bcr_class.csv')

cloupeout5 = left_join(cloupein, sce_cloupe) %>% select(Barcode, shm_rate)
write_csv(cloupeout5, 'refined/filtered_cluster_shm.csv')
```


## Subset clusters

```{r}
clusters_use = c('LMNA+', 'Memory', 'NR4A+', 'Naive(i)', 'Naive(ii)')
sce_cd20 = sce[,colData(sce)$ident %in% clusters_use]# & colData(sce)$pop != 'BLD']

```

## HVG

```{r}
blacklist = str_detect(rowData(sce_cd20)$symbol, 'IG[HKL]V[0-9]+')
```

```{r trendvar}
fit.g = scran::trendVar(sce_cd20, subset.row = !blacklist, use.spikes = FALSE, parametric = TRUE)
dec = scran::decomposeVar(sce_cd20, fit.g) %>% as.data.frame() %>% mutate(rank = rank(-bio/total))
ggplot(dec, aes(x = rank, y = p.value)) + geom_point()
rowData(sce_cd20) = cbind(dec, rowData(sce_cd20))

trendvar_tidy = rowData(sce_cd20) %>% as.data.frame %>% mutate(use = !blacklist, trend = fit.g$trend(mean), cut_mean = cut(sqrt(mean), 5)) %>% group_by(cut_mean, use) %>% mutate(rank_by_mean = rank(-bio/total))

ggplot(dplyr::filter(trendvar_tidy, use), aes(x = sqrt(mean), y = total))+geom_point() + geom_line(aes(y = trend), color = 'red') + ggrepel::geom_text_repel(aes(label = ifelse(rank_by_mean<10 & FDR < .1, symbol, '')), size = 2) + theme_minimal()
```

```{r hvg}
rowData(sce_cd20)$hvg = dec$rank<1000 | !is.na(rowData(sce_cd20)$curated)
```


# PCA

## Selected Seurat clusters

```{r pca}
PCA_COMPONENTS = 2
sce_cd20 = runPCA(sce_cd20, ntop = Inf, ncomponents = PCA_COMPONENTS, scale_features = TRUE, feature_set = rowData(sce_cd20)$hvg)

pca_obj = prcomp(t(logcounts(sce_cd20[rowData(sce_cd20)$hvg,])), rank = 10, center = TRUE, scale. = TRUE)
px = predict(pca_obj, t(logcounts(sce[rowData(sce_cd20)$hvg,])))
reducedDims(sce) = SimpleList(PCA = px)

sce_cd20 = runDiffusionMap(sce_cd20, ncomponents = 2, ntop = Inf, scale_features = TRUE, feature_set = rowData(sce_cd20)$hvg)



#plotPCA(sce_cd20, ncomponents = min(PCA_COMPONENTS, 4), colour_by = 'dataset')
#plotPCA(sce_cd20, ncomponents = min(PCA_COMPONENTS, 4), colour_by = 'total_features')

thresh_x = -3
thresh_y = 0
```

```{r}
plotPCA(sce_cd20, ncomponents = 2, colour_by = 'ident')
plotPCA(sce_cd20, ncomponents = 2, colour_by = 'ident') + geom_hline(yintercept = thresh_y, lty = 2) + geom_vline(xintercept = thresh_x, lty = 2)
plotPCA(sce_cd20, ncomponents = 2, colour_by = 'IGHD', alpha = .3)
plotPCA(sce_cd20, ncomponents = 2, colour_by = 'IGHDhi')

```

PCA, Selected groups.  PC1 = NR4A+ score, PC2 = memory score?

```{r}
plotPCA(sce_cd20, ncomponents = 2, colour_by = 'shm_rate')

plotPCA(sce_cd20[,!is.na(sce_cd20$shm_rate)], ncomponents = 2, colour_by = 'shm_rate')
```

SHM, and SHM in cells with a recovered BCR

## Diffusion maps, selected Seurat clusters

```{r}
plotDiffusionMap(sce_cd20, colour_by = 'ident')
plotDiffusionMap(sce_cd20, colour_by = 'grp')

```

Diffusion maps, selected groups


```{r}
plt = plotReducedDim(sce_cd20, use_dimred = 'PCA', ncomponents = 2, colour_by = 'dataset') + geom_point(aes(color = colour_by))

plt + facet_wrap(~colour_by)

plt = plotReducedDim(sce, use_dimred = 'PCA', ncomponents = 2, colour_by = 'ident') + geom_point(aes(color = colour_by))

plt + facet_wrap(~colour_by)



var = cumsum(attr(reducedDim(sce_cd20, 'PCA'), 'percentVar'))
qplot(x = seq_along(var), y = var) + geom_line() + scale_x_log10() + ylab('Proportion of variance') + xlab('Index') + theme_minimal()
```


## All groups 

```{r pca_allgroups}
sce = sce[,!is.na(sce$ident)]
plotPCA(sce, ncomponents = min(PCA_COMPONENTS, 4), colour_by = 'ident')

plotPCA(sce[,stringr::str_detect(sce$ident, 'Naive|LMNA|NR4A|Memory')], ncomponents = min(PCA_COMPONENTS, 4), colour_by = 'ident')
```

PCA (All groups)




# Slingshot

```{r}
use = 'PCA'
sce_cd20 = slingshot(sce_cd20, clusterLabels = 'ident', reducedDim = use)

# hist(sce_cd20$slingPseudotime_1)

colors <- viridis::plasma(100)
plot(reducedDim(sce_cd20, type = use), col = colors[cut(sce_cd20$slingPseudotime_1,breaks=100)], pch=16, asp = 1)
lines(SlingshotDataSet(sce_cd20), lwd=2)
```

Slingshot pseudotime places NR4A+ as an endpoint, which doesn't seem plausible.

```{r}
plot(reducedDim(sce_cd20, type = use), col = sce_cd20$res.0.5 , pch=16, asp = 1)
lines(SlingshotDataSet(sce_cd20), lwd=2)

pairs(cbind(sce_cd20$slingPseudotime_1, reducedDim(sce_cd20, type = use)[,1:2]), col = sce_cd20$grp)
```

```{r, eval = FALSE}
pt_regression = glm(sce_cd20$grp != 'TBX21-.ITGAX-' ~ sce_cd20$slingPseudotime_1, family = 'binomial')
dr_regression = glm(sce_cd20$grp != 'TBX21-.ITGAX-' ~ reducedDim(sce_cd20, use)[,1], family = 'binomial')
```

## PC1 heatmaps

```{r}
library(ComplexHeatmap)

sce_cd20plus = sce[, stringr::str_detect(sce$ident, 'Naive|LMNA|NR4A|Memory')]

scater_ident = plotPCA(sce_cd20plus, ncomponents = min(PCA_COMPONENTS, 4), colour_by = 'ident') 


scater_ident_b = ggplot_build(scater_ident)
scater_col = bind_cols(scater_ident$data, scater_ident_b$data[[1]]["fill"]) %>% 
  dplyr::select(fill, ident = colour_by) %>% unique()
scater_col_vec = setNames(scater_col$fill, scater_col$ident)

#prc = sce_cd20[rowData(sce_cd20)$hvg,] %>% assay('logcounts') %>% t %>% prcomp(scale. = TRUE, rank. = 2)

loading = pca_obj$rotation %>% as_tibble(rownames = 'symbol') %>% mutate_at(vars(rank1 = PC1, rank2 = PC2), function(x)  rank(desc(abs(x)))) %>% right_join(rowData(sce_cd20plus) %>% as.data.frame)

rowData(sce_cd20plus) = cbind(rowData(sce_cd20plus), loading[, c('PC1', 'PC2', 'rank1', 
'rank2')])
colData(sce_cd20plus) = cbind(colData(sce_cd20plus), reducedDim(sce_cd20plus)[,1:2])

genes_use = dplyr::filter(loading, rank1 <= 40) %>% arrange(PC1) %>% pull(symbol)

reorder_sce = function(sce_, genes_use, ordering_vector){
  sce_[genes_use, order(ordering_vector) ]
}

sce_cd20_pc1 = reorder_sce(sce_cd20plus, genes_use, sce_cd20plus$PC1)
sce_cd20_pc2 = reorder_sce(sce_cd20plus, dplyr::filter(loading, rank2 <= 40)$symbol,sce_cd20plus$PC2)
sce_cd20_curated = reorder_sce(sce_cd20plus, dplyr::filter(loading, !is.na(curated))$symbol,sce_cd20plus$PC1)



# heatmapf = function(SCE){
# SCE  %>% assay('logcounts') %>% as.matrix %>% Heatmap(cluster_columns = FALSE, clustering_distance_rows = 'pearson', top_annotation = HeatmapAnnotation(colData(SCE)[, c('dataset', 'ident')] %>% setNames(nm = c('dataset', 'Cluster')) , col = list(Cluster = scater_col_vec)), name = 'Normalized Expression')
# }

heatmapf = function(SCE, extra){
SCE  %>% assay('logcounts') %>% as.matrix %>% Heatmap( name = 'Normalized Expression', cluster_columns = FALSE, clustering_distance_rows = 'pearson', top_annotation = HeatmapAnnotation(colData(SCE)[, c('ident', extra), drop = FALSE] %>% setNames(nm = c('Cluster', extra)) , col = list(Cluster = scater_col_vec)))
}

```

```{r pc1_hm}
sce_cd20_pc1 %>% heatmapf(extra = 'PC1')
sce_cd20_pc1[,sce_cd20_pc1$PC2 < thresh_y & sce_cd20_pc1$ident %in% c('Naive(i)', 'Naive(ii)', 'NR4A+', 'Naive(iii)')] %>% heatmapf(extra = 'PC1')
```

Top loading genes and PC1

```{r pc1_shm}
plotColData(sce_cd20_pc1, x = 'PC1', y = 'shm_rate', colour_by = 'ident') + geom_smooth(method = 'lm', formula = y~splines::ns(x, df = 3)) + ylab('SHM Rate')

plotExpression(sce_cd20_pc1[,sce_cd20_pc1$PC2<thresh_y & sce_cd20_pc1$ident %in% c('Naive(i)', 'Naive(ii)', 'NR4A+', 'Naive(iii)')], x = 'PC1', features = c('NR4A1', 'DUSP1', 'TXNIP', 'CD79B'), colour_by = 'ident') + geom_smooth(method = 'lm', formula = y~splines::ns(x, df = 3)) + scale_fill_manual(name = NULL, values = scater_col_vec)
```

Correlation between shm and PC1

## PC2 heatmaps

```{r pc2_hm}
sce_cd20_pc2 %>% heatmapf(extra = 'PC2')
sce_cd20_pc2[,sce_cd20_pc2$PC1 > thresh_x & sce_cd20_pc2$ident %in% c('NR4A+', 'LMNA+', 'Memory', 'Naive(iii)')] %>% heatmapf(extra = 'PC2')
```

Top loading genes and PC2

```{r pc2_shm}
plotColData(sce_cd20_pc2, x = 'PC2', y = 'shm_rate', colour_by = 'ident') + geom_smooth(method = 'lm', formula = y~splines::ns(x, df = 3))  + ylab('SHM Rate')

plotExpression(sce_cd20_pc2[,sce_cd20_pc2$PC1>thresh_x & sce_cd20_pc2$ident %in% c('NR4A+', 'LMNA+', 'Memory', 'Naive(iii)')], x = 'PC2', features = c('JUN', 'DUSP1', 'S100A6', 'HOPX'), colour_by = 'ident') + geom_smooth(method = 'lm', formula = y~splines::ns(x, df = 3))  + scale_fill_manual(name = NULL, values = scater_col_vec)

plotExpression(sce_cd20_pc2[,sce_cd20_pc2$PC1>thresh_x], x = 'PC2', features = c('JUN', 'DUSP1', 'S100A6', 'HOPX'), colour_by = 'ident') + geom_smooth(method = 'lm', formula = y~splines::ns(x, df = 3))  + scale_fill_manual(name = NULL, values = scater_col_vec)
```

Correlation between SHM and PC2

## Curated heatmaps (ordered by PC1)

```{r}
sce_cd20_curated %>% heatmapf(extra = 'PC1')
```

# Seurat PCs

```{r, eval = FALSE}
library(Seurat)
foo = assays(sce_cd20) %>% lapply(as.matrix)
colnames(foo$counts) = paste(colData(sce_cd20)$dataset, colData(sce_cd20)$barcode, sep = '_')
#ssce = sce
#assays(ssce) = foo
#ssce = Seurat::Convert(ssce, to = 'seurat')
meta = colData(sce_cd20) %>% as.data.frame
rownames(meta) = colnames(foo$counts)
ssce = Seurat::CreateSeuratObject(raw.data = foo$counts, meta.data = meta)
ssce = Seurat::NormalizeData(ssce)
ssce@var.genes = rowData(sce_cd20) %>% as.data.frame %>% dplyr::filter(hvg) %>% pull(symbol)

```

```{r seurat_pca,  eval = FALSE}
PCA_COMPONENTS = 6
MAX_PCA = ceiling(2*PCA_COMPONENTS)

ssce = Seurat::ScaleData(ssce, display.progress = FALSE)
ssce = Seurat::RunPCA(ssce, pcs.print = 0, pcs.compute = MAX_PCA)
Seurat::PCElbowPlot(ssce, num.pc = MAX_PCA)

#ssce =  Seurat::JackStraw(object = ssce, display.progress = TRUE, num.replicate = 100, do.par = TRUE, num.cores = 3, maxit = 500, num.pc = MAX_PCA)
#Seurat::JackStrawPlot(ssce, PCs = 1:PCA_COMPONENTS)
Seurat::DimPlot(ssce, group.by = 'sample')
Seurat::DimPlot(ssce, group.by = 'pop')
Seurat::DimPlot(ssce, group.by = 'grp')



Seurat::PCHeatmap(ssce, pc.use = c(1:2), use.scale = FALSE)
Seurat::VizPCA(object = ssce, pcs.use = c(1:2))
```

```{r}
pc_table = cbind(colData(sce)[, c('sample', 'pop', 'Barcode')], reducedDims(sce)$PCA) %>%
  as.data.frame()
write_csv(pc_table, path = '05slingshot_files/principal_comp.csv')
```

# Tbet analysis (old)


```{r, results = 'asis'}
knitr::kable(table(dataset = colData(sce_cd20)$dataset, colData(sce_cd20)$grp) )
```

```{r runTests, eval = FALSE}
library(MAST)
options(mc.cores = 4)
sca = as(sce_cd20, 'SingleCellAssay')
rowData(sca)$primerid = rowData(sca)$Symbol
assays(sca) = list(logCounts = assay(sca, 'logcounts'))
zz_umi = zlm(~I(ITGAX>0)*I(TBX21>0) + total_features, sca = sca, method = 'bayesglm', parallel = TRUE)

zz = zlm(~I(ITGAX>0)*I(TBX21>0), sca = sca, method = 'bayesglm', parallel = TRUE)

dt = summary(zz_umi, doLRT = c('I(ITGAX > 0)TRUE:I(TBX21 > 0)TRUE', 'I(ITGAX > 0)TRUE', 'I(TBX21 > 0)TRUE'))$datatable %>% as_tibble()

```


```{r showTests, eval = FALSE}
contrasts = dt %>% dplyr::filter(str_detect(contrast, 'ITGAX|TBX'), component == 'H') %>% mutate(fdr = p.adjust(`Pr(>Chisq)`, method = 'fdr'))

sig = dplyr::filter(contrasts, fdr < .1, primerid != 'TBX21', primerid != 'ITGAX')


with_hurdle = dt %>% dplyr::filter(component == 'D') %>% right_join( (contrasts %>% semi_join(sig %>% select(primerid))) %>% select(primerid, contrast, fdr))

with_hurdle = with_hurdle %>% group_by(contrast) %>% mutate(primerid_rank = rank(fdr))

ggplot(with_hurdle, aes(x = coef, y = -log10(fdr), label = ifelse(primerid_rank < 30, primerid, '')))+ geom_point(size = 1) + theme_minimal() + facet_wrap(~contrast) + geom_text_repel(size = 2, segment.alpha = .25) + scale_colour_brewer(type = 'qual', palette = 2) + xlab('logFC(Group:Baseline)')

```

```{r, eval = FALSE}
sig %>% group_by(contrast) %>% summarize(n())
```


# CD20 tests
```{r, eval = FALSE}
colData(sca)$MS4A1 = assay(sca)['MS4A1',]

zz_msa = zlm(~MS4A1 + total_features, sca = sca[!is.na(rowData(sca)$group),], method = 'bayesglm', parallel = TRUE)
dt_msa = summary(zz_msa, doLRT = c('MS4A1'))$datatable %>% as_tibble()

ggplot(dplyr::filter(dt_msa, component == 'H', contrast == 'MS4A1'), aes(y = fct_reorder(primerid, `Pr(>Chisq)`), x = AMmisc::clamp(-log10(`Pr(>Chisq)`), 20))) + geom_point() + ylab('Gene') + xlab('-log10(P value)') + ggtitle('Correlates of CD20')

```

