---
title: "Phase1 / 10X Synovium Integration"
author: "Andrew McDavid"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
  slidy_presentation:
    font_adjustment: +1
extensions: inline_footnotes
---
  
  
```{r library_load}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, autodep=TRUE, message=FALSE, warning=FALSE)
knitr::opts_chunk$set(dev = c('png', 'pdf'))

library(SingleCellExperiment)
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(scater)

phase1 = readRDS('private/01_sce_bcell.rds')
phase1_sle = readRDS('private/01_sle_bcell.rds')
tenx = SingleCellExperiment(assay = readRDS('private/filtered_clustered_exprs.rds'),
                                            rowData = read_csv('private/filtered_clustered_fdata.csv'),
                                            colData = read_csv('private/filtered_clustered_cdata.csv'))
zheng = DropletUtils::read10xCounts('private/zhen_bcells_filtered_matrices_mex/hg19/')
```


## Day 0

```{r pitzalis_day0_br}
lazy_cellcall = function(sce_){
  br.out = DropletUtils::barcodeRanks(assay(sce_))
  plot(br.out$rank, br.out$total, log="xy", xlab="Rank", ylab="Total")
o = order(br.out$rank)
lines(br.out$rank[o], br.out$fitted[o], col="red")
abline(h=metadata(br.out)$knee, col="dodgerblue", lty=2)
abline(h=metadata(br.out)$inflection, col="forestgreen", lty=2)
legend("bottomleft", lty=2, col=c("dodgerblue", "forestgreen"), 
    legend=c("knee", "inflection"))
  br.out
}

call_intercalate = function(f, ..., extra){
  do.call(f, c(list(...), extra))
}

halfassed_qc = function(sce_, br.out, perCellQCArgs = list(percent_subsets = c('subsets_Mito_percent', 'percent_top_50'))) {
  ncell_start = ncol(sce_)
  is.mito = str_detect(rowData(sce_)$Symbol, '^MT-')
  df = perCellQCMetrics(sce_, subsets=list(Mito=is.mito))
  qqc = call_intercalate(quickPerCellQC, df, extra = perCellQCArgs)
  colData(sce_) = cbind(colData(sce_), df, qqc)
  
  qcplot = ggplot(colData(sce_) %>% as.data.frame(), aes(x = discard, y = sum)) + geom_boxplot() 
  print(qcplot + scale_y_log10())
  print(qcplot + aes(y = detected) + scale_y_log10())
  print(qcplot + aes(y = subsets_Mito_percent))
  print(qcplot + aes(y = percent_top_50))
  
  if(!missing(br.out)){
    empty = br.out$total < metadata(br.out)$knee
    sce_ = sce_[,!empty & !qqc$discard]
  } else{
    empty = rep(FALSE, ncol(sce_))
    sce_ = sce_[,!qqc$discard]
  }
  qc_table = data.frame(condition = c('start', 'empty', 'qc_fail', 'end'), ncells = c(ncell_start, sum(empty), sum(qqc$discard), ncol(sce_)))
  sce_ = logNormCounts(sce_)
  metadata(colData(sce_)) = list(qc = qc_table)
  sce_
}

```

# Combine data

```{r intersect}
# Disable log file cluttering 
futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")

vd = VennDiagram::venn.diagram(list(phase1_ra = rownames(phase1), 
                                    tenx = rownames(tenx), 
                                    phase1_sle = rownames(phase1_sle),
                                    zheng = rowData(zheng)$Symbol),
                                    filename = NULL, fill = c('red', 'yellow', 'blue', 'grey50'), alpha = rep(.35, 4))
grid::grid.draw(vd)

common_id = purrr::reduce(list(rownames(phase1), rownames(tenx), rownames(phase1_sle), rowData(zheng)$Symbol), intersect)
```


```{r}
phase1_inter = phase1[common_id,]
tenx_inter = tenx[common_id,]
cd_phase1_inter = colData(phase1_inter) %>% as.data.frame() %>% select(Barcode = cell_name, orig_bc = barcode, sample, type, cluster, disease) %>% mutate(run = 'phase1')

cd_tenx_inter = colData(tenx_inter) %>% as.data.frame() %>% select(subject = sample, sample = dataset, Barcode, chain_type,  class_heavy, class_light, shm_rate, res.0.5, ident) %>% mutate(cluster = str_c('TX', res.0.5), Barcode = make.unique(Barcode), run = 'tenx')

rd_phase1_inter = rowData(phase1_inter) %>% as.data.frame() %>% select(symbol, group)
rd_phase1_tenx =  rowData(tenx_inter) %>% as.data.frame() %>% select(symbol = Symbol)

coldat_inter = bind_rows(cd_phase1_inter, cd_tenx_inter)
rd_inter = left_join(rd_phase1_inter, rd_phase1_tenx)

sce = SingleCellExperiment(assay = cbind(Matrix::Matrix(counts(phase1_inter), sparse = TRUE), assay(tenx_inter)),
                           rowData = rd_inter, colData = coldat_inter)
colnames(sce) = sce$Barcode
assayNames(sce) = 'counts'
blacklist = str_detect(rowData(sce)$symbol, 'IG[HKL]V[0-9]+')
blacklist_name = rowData(sce)$symbol[blacklist]
```


## Zheng QC

```{r}
br.out = lazy_cellcall(zheng)
```

```{r zhengqc}
zheng = halfassed_qc(zheng, br.out)
knitr::kable(metadata(colData(zheng))$qc)

```


```{r zhengfilt}
zheng = zheng[!duplicated(rowData(zheng)$Symbol),]
rownames(zheng) = rowData(zheng)$Symbol
plotExpression(zheng, x = 'MS4A1', features = c('CD3E', 'CD14', 'CD19', 'XBP1', 'TCL1A', 'HOPX'))
```

No expression of CD3 or CD14, otherwise seems reasonable?

## Joint QC

```{r jqc}
sce = halfassed_qc(sce, perCellQCArgs = list(percent_subsets = NULL, batch = factor(sce$run)))
knitr::kable(metadata(colData(sce))$qc)
saveRDS(sce, file = 'refined/amp_sle_ra_sce.rds')
```

```{r monocyte_qc}
sce$quickclust = scran::quickCluster(sce)
plotExpression(sce, x = "PDPN", features = c("FN1", 'COL3A1'), colour_by = "run", exprs_values = 'logcounts')

plotExpression(sce, x = 'CD14', features = 'CXCL16', colour_by = 'quickclust')
plotExpression(sce, x = 'quickclust', features = c('CD14', 'CXCL16'), colour_by = 'run')
```



# SLE Phase 1

```{r}
phase1_sle = logNormCounts(phase1_sle)

ggplot(as.data.frame(colData(phase1_sle)), aes(x = sample, fill = cluster)) + geom_bar() + coord_flip()

```

# SingleR

## Train on 10x, predict on AMP and Zheng

```{r singler}
ref = sce[,sce$run == 'tenx' & !is.na(sce$ident)]

train = SingleR::trainSingleR(ref = ref, labels = ref$ident, de.method = "classic")
phase1 = sce[,sce$run == 'phase1']

pred_tenx = SingleR::classifySingleR(ref, train)
pred_phase1 = SingleR::classifySingleR(phase1, train)
pred_sle = SingleR::classifySingleR(phase1_sle, train)
pred_zheng = SingleR::classifySingleR(zheng, train)
```

```r

merge.SingleCellExperiment = function(x, y, ...){
    nx = x
   rd = getS3method('merge', class(rowData(x)))(x, y, ...)
   cd = colData(x)
}

```


## Classifying our 10x data

```{r singler_10x}
SingleR::plotScoreHeatmap(pred_tenx, clusters = ref$ident, show.labels = FALSE)
```

## Classifying phase 1

```{r singler_ampra}
SingleR::plotScoreHeatmap(pred_phase1, clusters = phase1$cluster)
SingleR::plotScoreDistribution(pred_phase1, show = "delta.med", show.nmads = 3)

```

Shows scores of classified vs non-classified cells.  Horizontal line is the lines for rejecting any classification.

## Classifying SLE phase1

```{r singler_ampsle}
SingleR::plotScoreHeatmap(pred_sle, clusters = phase1_sle$cluster)
SingleR::plotScoreDistribution(pred_sle, show = "delta.med", show.nmads = 3)


pred_phase1_sle_df = cbind(pred_sle, colData(phase1_sle))

ggplot(as.data.frame(pred_phase1_sle_df), aes(x = sample, fill = pruned.labels)) + 
  geom_bar() + coord_flip()

```

## Classifying Zheng CD19 B

```{r singler_zheng}
SingleR::plotScoreHeatmap(pred_zheng)
SingleR::plotScoreDistribution(pred_zheng, show = "delta.med", show.nmads = 3)


pred_zhang_df = cbind(pred_zheng, colData(zheng), sample = 'Zheng PBMC', cluster = 'CD19+ B')
```



```{r singler_phase1_distr}
pred_phase1_df = cbind(pred_phase1, colData(phase1))

sample_prop = as.data.frame(pred_phase1_df) %>% group_by(pruned.labels, sample, disease) %>% 
  summarize(n_cluster = n()) %>% group_by(sample, disease) %>% mutate(n_overall = sum(n_cluster),
                                                                      prop = n_cluster/n_overall)


crosstab = as.data.frame(xtabs(~pruned.labels + cluster + sample, data = as.data.frame(pred_phase1_df), addNA = TRUE))

crosstab2 = as.data.frame(xtabs(Freq ~sample, data = crosstab)) %>% rename(total = Freq)
crosstab = left_join(crosstab, crosstab2) %>% mutate(prop = Freq/total)


#crosstab = as.data.frame(pred_phase1b) %>% group_by(pruned.labels, cluster, sample, disease) %>% 
#  summarize(n_cluster = n()) %>% group_by(sample, disease) %>% mutate(n_overall = sum(n_cluster),
#                                                                      prop = n_cluster/n_overall) %>% #tidyr::complete(pruned.labels, cluster, tidyr::nesting(sample, disease), fill = list(prop = 0))

ggplot(crosstab, aes(x = cluster, y = pruned.labels, fill = prop)) + geom_tile() +
  facet_wrap(~sample) + scale_fill_distiller()

ggplot(sample_prop, aes(x = pruned.labels, y = prop)) + geom_boxplot()
ggplot(sample_prop, aes(x = sample, fill = pruned.labels, y = prop)) + 
  geom_col() + coord_flip()

```


```{r singler_allexp}
pred_tenx = cbind(pred_tenx, colData(ref))
all_df = list(synovium = pred_tenx[, c('scores', 'pruned.labels', 'cluster', 'sample')],
  amp_sle = pred_phase1_sle_df[, c('scores', 'pruned.labels', 'cluster', 'sample')], 
                 amp_ra = pred_phase1_df[, c('scores', 'pruned.labels', 'cluster', 'sample')],
              peripheral_b = pred_zhang_df[, c('scores', 'pruned.labels', 'cluster', 'sample')]
  )

all_pred = purrr::map_dfr(all_df, as.data.frame, .id = 'experiment')

all_pred = mutate(all_pred, experiment = ifelse(sample == 'RA134_BLD', 'peripheral_b', experiment))

sample_exp_tab = unique(select(all_pred, experiment, sample)) %>%
  mutate(Experiment = factor(experiment, levels = c('amp_sle', 'peripheral_b','amp_ra', 'synovium'), labels = c('SLE (AMP)',  'Peripheral B', 'Synovium (AMP)', 'Synovium (10x)')))

score_plot = ggplot(all_pred, aes(x = cluster, y = scores.NR4A.))  + 
  facet_grid(~experiment, space = 'free_x', scales = 'free_x') + ylab('Spearman Correlation NR4A+')

score_plot + ggbeeswarm::geom_quasirandom()
score_plot + geom_boxplot()
score_plot + aes(y = scores.Plasma.i.) + geom_boxplot()
score_plot + aes(x = pruned.labels) + geom_boxplot()

ggplot(all_pred, aes(x = sample, fill = pruned.labels)) + geom_bar() + 
  facet_wrap(~experiment, scales = 'free') + coord_flip()
summary(lm(scores.NR4A. ~ experiment, data = filter(all_pred, pruned.labels == 'NR4A+')))

```

Compare 28 NRA41+ cells in SLE to 835 cells in RA.
Maybe also compare the gap between first and second or first and third (since second is likely LMNA+) best cluster?

```{r}
sample_tab = table(sample = all_pred$sample, 'NR4A+' = all_pred$pruned.labels== 'NR4A+')

knitr::kable(sample_tab)
```

We'll keep subjects with >=5 cells.

```{r nr4_freq}
sample_tab = sample_tab[rowSums(sample_tab) >= 5,]

sample_tib = as_tibble(unclass(sample_tab), rownames = 'sample') %>% left_join(sample_exp_tab)

nr4_lme = lme4::glmer(cbind(`TRUE`, `FALSE`) ~ experiment + (1|sample), family = 'binomial', data = sample_tib)
exp_tab = data.frame(experiment = unique(sample_tib$experiment, sample = NA))

bs = lme4::bootMer(nr4_lme, nsim = 100, re.form = NA, FUN = function(m) predict(m, newdata =exp_tab , re.form = NA, type = 'response'))

nr4_fitted = cbind(estimate = bs$t0, exp_tab, confint(bs)) %>% right_join(unique(sample_exp_tab[c('experiment', 'Experiment')]))


nr4_tab = as.data.frame(prop.table(sample_tab, margin = 1)) %>% left_join(sample_exp_tab) %>% filter(NR4A. == TRUE)

ggplot(nr4_tab, aes(x = Experiment, y = Freq)) + ggbeeswarm::geom_beeswarm() + geom_crossbar(data = nr4_fitted, aes(y = estimate, ymin = `2.5 %`, ymax = `97.5 %`), color = 'blue', width = .1) + theme_minimal() + ylab("NR4A+ Frequency")
```

```{r}
summary(nr4_lme)
```

# Correlate AMP-1 with singleR labels

```{r}
cluster_id = readRDS('private/celseq_synovium_meta_5265cells_paper.rds')

cluster_freq = cluster_id %>% group_by(cell_type) %>% do({
  cluster = factor(.$cluster)
  sample = factor(.$sample)
  tab = table(cluster, sample)
  as.data.frame(tab)
}) %>% group_by(cell_type, sample) %>% mutate(total = sum(Freq), percent = Freq/total*100)

phase1_join = inner_join(sample_tib, cluster_freq) 


```

```{r nr4a_phase1_cluster, dev = c('png', 'pdf')}
#sce_t = readRDS('01_sce_tcell.rds')

#t_pop = colData(sce_t) %>% as.data.frame() %>% filter(!is.na(cluster))

# count_pops = . %>% group_by(sample, disease, cluster) %>% summarize(n()) %>% tidyr::complete(cluster, tidyr::nesting(sample, disease), fill = list(`n()` =0)) %>%
#   group_by(sample, disease) %>% mutate(total = sum(`n()`), prop = `n()`/total) 
# 
# t_pop = t_pop %>% count_pops 
# 
# b_pop = md %>% count_pops
# 
# both_pop = inner_join(t_pop, b_pop, by = c('sample', 'disease'), suffix = c('_t', '_b'))

pop_tests = phase1_join %>% group_by(cluster) %>% do({
  dat = .
  gg = lme4::glmer(cbind(`TRUE`, `FALSE`) ~ percent+ (1|sample),  data = dat, family= 'binomial')
  broom::tidy(gg, conf.int = TRUE)
})

ptests = filter(pop_tests, term == 'percent') %>% ungroup() %>% mutate(fdr = p.adjust(p.value)) %>% select(-term) %>% rename(term = cluster)

dotwhisker::dwplot(ptests) + geom_text(aes(x = estimate, y = term, label = ifelse(fdr < .01, '<.01', round(fdr, 2))), position = position_nudge(y = .2, x = .04)) + ylab("Phase 1 cluster") + xlab("Log odds of NR4A+\n per % phase 1")

knitr::kable(ptests)
```
