---
title: "QC and Differential Expression for amp_bulk project"
subtitle: "Template version 0.0.0.9001"
author: "Andrew McDavid"
date: '`r Sys.Date()`'
params:
  input_root: "refined/01_load_se.rds"
  output_root: "refined/02_qc_dea" #prefix for output
  se_subset: NULL # NULL or character. An arbitrary R expression to be parsed and evaled.
                   # should assign to `se`.
  plot_covariates: NULL #character vector in colData(se) used as aestetics for some plots
  sample_id: 'sample_id' #field in colData(se) containing sampleID identifiers
  features_csv: NULL # csv with fields SYMBOL / gene_group.
  fdr_level: 0.1 # significance level used for plots / venn diagrams
  terms_annotation_csv: NULL # csv file containing a cross-walk between the models fit, terms r turned, the term_class (used to group equivalent terms in stratified models), plot_volcano ` ogical` if a volcano plot should be made
  design_csv: NULL # csv with fields name,formula,subset_expression,contrast.  Used to fit D Seq2 models.
  dea_method: "deseq2" # "deseq2", "nebula" or "glmer.nb"
  cache: FALSE
  organism: 'mouse'
output:
  rmdformats::html_clean:
    code_folding: hide
    toc: yes
    toc_float: yes
  github_document: default
---

Notes:
    
```{r libraries, results='hide', echo = FALSE, warning = FALSE, message = FALSE}
suppressPackageStartupMessages({
library(DESeq2)
library(dplyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(readr)
library(tibble)
library(tidyr)
library(biobroom)
library(broom)
library(ComplexHeatmap)
})
source("geneseebulk-functions.R")
knitr::opts_chunk$set(warning = FALSE, error = FALSE, message = FALSE, echo = TRUE, dev = c('png', 'pdf'), cache = params$cache, autodep = params$cache)
```

Knit with:
```{r}
print(params)
```


```{r load, cache.extra = params$input_root}
se = readRDS(params$input_root)
```

# Cross-type associations with NR4As

```{r setup-cross}
genes3 = tibble(SYMBOL = c('NR4A1', 'NR4A2', 'NR4A3', 'CD69', 'CD83', 'LMNA', 'GPR183', "LY9", "CXCR4"))

genes_nr4a = left_join(genes3, rowData(se) %>% as.data.frame() %>% dplyr::select(SYMBOL, ENSEMBL))
se_b2 = se[, se$Cell.type == 'B cell']
nr4a_total_score = colMeans(assay(se_b2, 'lcpm')[genes_nr4a$ENSEMBL,])
nr4a2 = assay(se_b2, 'lcpm')[filter(genes_nr4a, SYMBOL == 'NR4A2')$ENSEMBL,]
nr4a_covariates = data.frame(Donor.ID = se_b2$Donor.ID, nr4a_total_score = scale(nr4a_total_score), nr4a2 = scale(nr4a2))

ggplot(nr4a_covariates, aes(x = nr4a_total_score, y = nr4a2, label = Donor.ID)) + geom_text()

colData(se) = left_join(as.data.frame(colData(se)), nr4a_covariates) %>% DataFrame()
```


```{r subset, eval = !is.null(params$se_subset), code = params$se_subset}

```

```{r}
designs = read_csv(params$design_csv, comment = "#")
```

```{r}
if(ncol(colData(se))<1) stop("Expecting non-empty colData")
if(is.null(params$plot_covariates)){
  plot_covariates = rep(names(colData(se)), 2)[1:2] # ensure at least length two
} else{
  plot_covariates = params$plot_covariates
}
```

```{r}
dge = DESeqDataSet(se=se, design = as.formula(designs$formula[1]))
```



```{r}
#design(dge) = ~ groupf

#genes_plus_random = union(which(!is.na(rd$group)), sample(nrow(rd), 200))

logcpm = function(x){
    total = colSums(counts(x))
 log2( 1e6*t(counts(x))/total + 1)
}

colData(dge) = cbind(colData(dge), logcpm(dge)[, !is.na(rowData(dge)$group)])
```





```{r filter}
dge_f = dge#[!duplicated(dge$genes$ENSEMBL) | is.na(dge$genes$ENSEMBL),]
#dge_f = dge[,dge$group == params$group]
dge_f = estimateSizeFactors(dge_f)
null = rowSums(counts(dge_f) > 5) < 3
dge_f = dge_f[!null,]
ds_trans = vst(dge_f)
```


# Differential expression testing

Currently we test `r nrow(designs)` different models.  Specifications are shown below:

```{r}
knitr::kable(designs)
```

```{r}
dea_method = match.arg(params$dea_method, c('deseq2', 'nebula'))
```


```{r run-deseq2, eval = dea_method == 'deseq2'}
dew = purrr::map2(setNames(designs$formula, designs$name), designs$subset_expr, function(x, y) run_deseq_design(dge_f,  as.formula(x), y))
```



```{r}
dew_df = purrr::map_dfr(dew, tidy, .id = 'model')  %>% dplyr::rename(SYMBOL = gene, padj = p.adjusted, pvalue = p.value, log2FoldChange = estimate)
models_and_terms = dew_df %>% count(model, term)
knitr::kable(models_and_terms)
```

Models fit and terms returned.

```{r, eval = !is.null(params$terms_annotation_csv)}
models = read_csv(params$terms_annotation_csv)
```

```{r}
 
count_sig = semi_join(dew_df, models) %>% group_by(SYMBOL) %>% 
    summarize(n_sig = sum(padj < .1, na.rm = TRUE), geomean_pvalue = sum(log10(pvalue), na.rm = TRUE), geovar_pvalue = var(log10(pvalue), na.rm = TRUE))

sig_contrasts = semi_join(dew_df, count_sig %>% filter(rank(geomean_pvalue)<20)) %>% left_join(models)

plt = ggplot(sig_contrasts, aes(y = SYMBOL, x = log2FoldChange, color = model)) + geom_text(aes(label = term_class))
plt + geom_vline(xintercept = 0, lty = 2)
```



```{r heatmaps, fig.height = 7}
models_plot = filter(models, plot_volcano)

for(m in seq_len(nrow(models_plot))){
  this_model = models_plot[[m,'model']]
  print(modified_volc(dew[[this_model]],  name = models_plot[[m, 'term']], heatmap_top_group = plot_covariates, heatmap_max_gene = 40,
                      title = sprintf("%s in %s background", models_plot[[m, 'term_class']], this_model)))
                               # ^^^ Modify as needed
}
```

## Venn diagrams of results

```{r venn}
terms_by_model = left_join(models, dew_df) %>% ungroup() %>% 
  filter(padj < params$fdr_level) %>% mutate(direction = sign(log2FoldChange))

terms_by_model_list = terms_by_model %>% split(.$term_class)

for(i in seq_along(terms_by_model_list)){
  gene_subset = terms_by_model_list[[i]] %>% select(SYMBOL, model)
  n_model = count(gene_subset, model)
  gene_subset = left_join(gene_subset, n_model) %>% 
    transmute(SYMBOL, model = glue::glue("{model}({n} genes)"))
  v = venneuler::venneuler(gene_subset)
  
  plot(v, main = names(terms_by_model_list)[i])
}
```

Genes significant at `r params$fdr_level` by model group.


Venn diagram

# Gene set enrichment

```{r hyper-geo-setup}
library(clusterProfiler)

db = switch(params$organism[1],
            mouse = "org.Mm.eg.db",
            human = "org.Hs.eg.db")

remap = bitr(unique(dew_df$SYMBOL), 'ENSEMBL', 'ENTREZID', db, drop = FALSE)

# take subset of term, and create model below...
gsea_genes = left_join(dew_df, remap, by = c('SYMBOL' = 'ENSEMBL')) %>% 
  mutate(direction = factor(sign(log2FoldChange), labels = c('-1' = 'down', '1' = 'up'))) %>% 
  left_join(models) %>% 
  filter(padj < params$fdr_level, abs(log2FoldChange) > 1, plot_volcano) %>% 
  group_by(term, direction)

term_direction_drop = gsea_genes %>% 
  group_by(model, direction) %>%
  summarize(n = n()) %>% 
  filter(n < 10)
```


```{r hyper-geo-drop, results = 'as.is'}
knitr::kable(term_direction_drop)

gsea_genes = gsea_genes %>% anti_join(term_direction_drop)
```

Dropping the following sets for having < 10 enriched genes.

```{r hyper-geo}
cc_go = compareCluster(ENTREZID ~ model + direction, 
                       data = gsea_genes, 
                       OrgDb = db, readable = TRUE, universe = unique(remap$ENTREZID), 
                       fun = 'enrichGO')
```

```{r clusterGO, fig.width=6, fig.height=4}
dotplot(cc_go, font.size = 6, showCategory = 12) + 
  theme(axis.text.x = element_text(angle = 90))
```



```{r write_results, eval = FALSE}
all_go = rbind(cc_go@compareClusterResult,
      cc_bmt@compareClusterResult,
      cc_c7@compareClusterResult)

readr::write_csv(all_go, path = 'enrichment_by_subpop.csv')
readr::write_csv(dew_df, path = 'tests_by_subpop.csv')
```

```{r write_deseq}
to_write = left_join(dew_df, models_plot)
write_csv(to_write, paste0(params$output_root, '_results.csv'))
```
